Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-03 SCP2008/paper/metamonads.tex
more writeup

@@

Parameterizing by the kind of container representing a matrix is
almost as straightforward.  Our containers are  parametric
over a |DOMAIN|, i.e., functors from a |DOMAIN| module
to the actual implementation of a container. The functor signature
|CONTAINER2D| specifies that a container must provide functions |dim1|
and |dim2| to extract the dimensions, functions |get| and |set| to
generate container getter and setters, the cloning generator |copy|
and functions that generate code for row and column swapping. The
inclusion of these functions in the signature of all containers makes
it simpler to optimize the relevant functions depending on the actual
representation of the container while not burdening the users of
containers with efficiency details.

@@

\begin{code2}
(* The `keyword' list of all the present internal features *)
module type INTERNAL_FEATURES = sig
  module R      : TrackRank.RANK
  module P      : TRACKPIVOT
  module L      : LOWER
end

module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
  module IF : INTERNAL_FEATURES
  type res
  val make_result : 'a wmatrix ->
   (<classif : 'a; 
     state : [> `TDet of 'a OD.Det.lstate |
                'a IF.R.tag_lstate |
                `TPivot of 'a IF.P.lstate |
                `TLower of 'a IF.L.lstate]; ..>,res) cmonad
   (*
    ('a,res,[> 'a OD.Det.tag_lstate | 'a IF.R.tag_lstate 
             | 'a IF.P.tag_lstate   | 'a IF.L.tag_lstate],'w) cmonad
      *)
end
module OutDetRank(OD : OUTPUTDEP) = struct
  module IF = struct
      module R   = Rank
      module P   = DiscardPivot
      module L   = NoLower end
  type res = C.contr * C.Dom.v * int
  let make_result m = perform
    det  <-- OD.Det.fin ();
    rank <-- IF.R.fin ();
    ret (Tuple.tup3 m.matrix det rank)
  (* Initialization: check the preconditions of instantiation of this struct*)
  let _ = OD.Det.fin ()
  let _ = IF.R.fin ()
end
\end{code2}

>>>>

A more complex domain: |Zp|
\begin{code}
module ZpMake(P:sig val p:int end) = struct
    type v = int
    let kind = Domain_is_Field
    let zero = 0 and one = 1
    let plus x y = (x + y) mod P.p
    let times x y = (x * y) mod P.p
    \dots
    let normalizer = None and better_than = None
    let () = assert (is_prime P.p)
end
\end{code}

This domain is parametrized by a an integer number |p|, the
(modulus?). Thus, ZpMake is a functor. It creates a domain that is a
field, but with no defined order. Hence
|normalizer| and |better_than| are set to |None|. However, |Zp| forms
a field only when |p| is prime. Therefore, we must make this check,
see the last line of that code. The expression 
|assert (is_prime P.p)| is the initializing expression. It is executed
when the correspo0nding module is instantaited
\begin{code}
module Z19 = ZpMakeL(struct let p = 19 end)
\end{code}
if we replace |19| with |9| above, we receive an error. That is a
run-time error. It is raised however as we instantiate and
combine modules that will eventually make the generator. Thus,
although the error is reported at run-time rather than compile time as
one might have hoped, the error is raised when \emph{generating the
generator} -- well before the generation of the target code could
begin. In our code we make extensive use of these so-called preflight 
checks performed as part of module initialization. The checks seem to
offer a good compromise: they are dynamic and so do not require
complicated type systems; on the other hand, the checks are performed
quite early, when buiding code generators, and so ensure that no 
code violating the corresponding \emph{semantic} constraints 
will be generated. Although some may frown on the use of the module
initializing expressions (since that requires careful attention to the
sharing and problems of multiple instantiation of a module), these
concerns do not apply to our case: our preflight tests are all idempotent
and maintain no state.

Lifting:

@@

\begin{code}
module type CONTAINER2D = sig
  module Dom:DOMAINL
  type contr
  type 'a vc = ('a,contr) rep
  type 'a vo = ('a,Dom.v) rep
  val getL : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo
  val dim1 : 'a vc -> ('a,int) rep
  val dim2 : 'a vc -> ('a,int) rep
  val mapper : ('a vo -> 'a vo) option -> 'a vc -> 'a vc
  val copy : 'a vc -> 'a vc
  val init : ('a,int) rep -> ('a, int) rep -> 'a vc
  val augment : 'a vc -> ('a,int) rep -> ('a, int) rep -> 'a vc ->
                ('a, int) rep -> 'a vc
  val identity : ('a,int) rep -> ('a, int) rep -> 'a vc
  val swap_rows_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val swap_cols_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val row_head : 'a vc -> ('a, int) rep -> ('a, int) rep -> 'a vo
  val col_head_set : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo -> 
            ('a,unit) rep
end
\end{code}

Parameterizing by the kind of container representing a matrix is
almost as straightforward.  The type of our containers include the
lifted domain |Dom| as one of the component. Particular containers are
Our containers are  parametric
over a |DOMAINL|, i.e., functors from a |DOMAINL| module
to the actual implementation of a container. For example, the
following functor defines a matrix container represented as arrays of
rows.

\begin{code}
module GenericArrayContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x n m = .< (.~x).(.~n).(.~m) >.
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  \dots
\end{code}
%
We also have containers wheer elements are stored in a 1D array, in a
row-wise (C-like) or column-wise (Fortran-like) modes.

The signature |CONTAINER2D| specifies that a container must provide
functions |dim1| and |dim2| to extract the dimensions, functions |getL|
to generate container getters, the cloning
generator |copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details.

@@

\begin{code2}
module type OUTPUTDEP = sig 
    module PivotRep : PIVOTKIND 
    module Det      : DETERMINANT
end
(* The `keyword' list of all the present internal features *)
module type INTERNAL_FEATURES = sig
  module R      : TrackRank.RANK
  module P      : TRACKPIVOT
  module L      : LOWER
end

module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
  module IF : INTERNAL_FEATURES
  type res
  val make_result : 'a wmatrix -> (\dots,res) monad
end
module OutDetRank(OD : OUTPUTDEP) = struct
  module IF = struct
      module R   = Rank
      module P   = DiscardPivot
      module L   = NoLower end
  type res = C.contr * C.Dom.v * int
  let make_result m = perform
    det  <-- OD.Det.fin ();
    rank <-- IF.R.fin ();
    ret (Tuple.tup3 m.matrix det rank)
  (* Initialization: check the preconditions of instantiation of this struct*)
  let _ = OD.Det.fin ()
  let _ = IF.R.fin ()
end
\end{code2}

We have several Rank aspects:
\begin{code2}
module Rank = struct
  include TrackRank
  let fin () = perform
      r <-- mo_lookup ip;
      ret (liftGet r)
end

module NoRank = struct
  include TrackRank
  let fin () = failwith "Rank is needed but is not computed"
end
\end{code2}

Both contain the |fin ()| function to generate the code representing
the computed rank. Only the |NoRank| does not track any rank, and so 
|fin ()| raises an error. The code 
|let _ = IF.R.fin ()| in |OutDetRank| invokes that fin function, which
will produce the monadic code generating action, or raise teh
error. We do not run the action at that time -- we only make sure
there is an action to run. This is the preflight check again, to rule
out the semantic error when the user specified that rank should be
computed and returned and yet specified the |NoRank| aspects which
computes no rank at all.
